Rotation effects on the Lyman-$\alpha$ line morphology in distant galaxies

### 2013-06-08 paper/Analisis.tex
Fig 1. fixed
updated results subsection on maxima

\newcommand{\LCDM}{$\Lambda$CDM~}
\newcommand{\beq}{\begin{eqnarray}}  
\newcommand{\eeq}{\end{eqnarray}}  
\newcommand{\zz}{$z\sim 3$} 
\newcommand{\apj}{ApJ}  
\newcommand{\apjs}{ApJS}  
\newcommand{\apjl}{ApJL}  
\newcommand{\aj}{AJ}  
\newcommand{\mnras}{MNRAS}  
\newcommand{\mnrassub}{MNRAS accepted}  
\newcommand{\aap}{A\&A}  
\newcommand{\aaps}{A\&AS}  
\newcommand{\araa}{ARA\&A}  
\newcommand{\nat}{Nature}  
\newcommand{\physrep}{PhR}
\newcommand{\pasp}{PASP}    
\newcommand{\pasj}{PASJ}    
\newcommand{\avg}[1]{\langle{#1}\rangle}  
\newcommand{\ly}{{\ifmmode{{\rm Ly}\alpha~}\else{Ly$\alpha$~}\fi}}
\newcommand{\hMpc}{{\ifmmode{h^{-1}{\rm Mpc}}\else{$h^{-1}$Mpc }\fi}}  
\newcommand{\hGpc}{{\ifmmode{h^{-1}{\rm Gpc}}\else{$h^{-1}$Gpc }\fi}}  
\newcommand{\hmpc}{{\ifmmode{h^{-1}{\rm Mpc}}\else{$h^{-1}$Mpc }\fi}}  
\newcommand{\hkpc}{{\ifmmode{h^{-1}{\rm kpc}}\else{$h^{-1}$kpc }\fi}}  
\newcommand{\hMsun}{{\ifmmode{h^{-1}{\rm {M_{\odot}}}}\else{$h^{-1}{\rm{M_{\odot}}}$}\fi}}  
\newcommand{\hmsun}{{\ifmmode{h^{-1}{\rm {M_{\odot}}}}\else{$h^{-1}{\rm{M_{\odot}}}$}\fi}}  
\newcommand{\Msun}{{\ifmmode{{\rm {M_{\odot}}}}\else{${\rm{M_{\odot}}}$}\fi}}  
\newcommand{\msun}{{\ifmmode{{\rm {M_{\odot}}}}\else{${\rm{M_{\odot}}}$}\fi}}  
\newcommand{\lya}{{Lyman$\alpha$~}}
\newcommand{\clara}{{\texttt{CLARA}}~}
\newcommand{\rand}{{\ifmmode{{\mathcal{R}}}\else{${\mathcal{R}}$ }\fi}}  
\newcommand{\hs}{{\hspace{1mm}}}  
\newcommand{\kms}{\,km~s$^{-1}$}

@@

@@

These results for these outgoing spectra are constructed by taking into
account all outgoing photons regardless of the direction of
propagation. In subsection \ref{subsec:direction} we report different
statistics aiming at quantifying the changes in the observed spectra
for observers with different viewing angles.

\begin{figure*}
  \includegraphics[width=0.45\textwidth]{7tDifSpeedsZ.png}
  \includegraphics[width=0.45\textwidth]{7tHOMDifSpeeds1.png}
\caption{Shape of the lyman alpha line for
    different velocities. The left (right) panel shows the central
    (homogeneous) photon distribution. All photons were taken into
    account regardless of their outgoing direction of propagation.
  \label{fig:differentvelocities}}
\end{figure*}

In the following subsections we quantify the trends observed in
Fig. \ref{fig:differentvelocities} and
Fig. \ref{fig:differentobservers} as a function of the maximum
rotation velocity $V_{\rm max}$ and the position of the observer with
respect to the rotation angle, $\mu = \cos{\theta}$. All the results
in this section will be expressed in terms of the adimensional
frequency $x$.

First we quantify the line by its full width at half maximum
(FWHM) and the peak positions. In order to interpret these results we
measure how the rotation affects the average number of scatterings
for each Lyman-alpha photon in the simulation. This help us to
introduce the next subsection on the escape fraction in our models
including dust. We conclude the section by estimating the
expected line flux for top hat filters at a fixed center and varying
width.

@@

\begin{figure}
    \includegraphics[width=0.45\textwidth]{WidthvsVmax.png}
\caption{Width of the lyman-alpha line for all the models.\label{fig:widthvsvelocity}}
\end{figure}

The first quantitative conclusion of the effect of rotation in the
Lyman alpha line is that the double peaks in the line tend to broaden
until they reach a single broad emission peak. This is most evident in
the case of Lyman-alpha sources homogeneously distributed in the gas
distributions (Fig.~\ref{fig:differentvelocities} Right panel).

To quantify the line broadening we measure a modified version of the full
width at half maximum (FWHM) for half of the line, $W_{1/2}$. It means that in
the case of double peaked emission, $W_{1/2}$ corresponds to the with
one of the peaks, while in the extreme case when the line is converted
into a single peak, $W_{1/2}$ corresponds to half of the full width.

This definition allows us to quantify the line width both in the cases
of double and single peak emission. Furthermore it has the advantage
that this line width should have a direct observational correspondence
to the observed line feature once the Inter-Galactic Medium (IGM)
effects are taken into account, which have the central effect of
strongly reducing the intensity of the blue peak of the line.

Fig.~\ref{fig:widthvsvelocity} summarizes our findings for $W_{1/2}$
as a function of $V_{\rm max}$. The line width increases with the
rotational velocity of the gas cloud. This increase can be of a factor
of $2-3$ with respect to the width with respect to the static
case. This trend is conserved at all optical depths regardless of the
initial source distribution.

%Im not sure about this anymore---TALK to Jaime about this!
This result includes all the outgoing photons, regardless of the
position of the observer. In Figure~\ref{fig:widthvstheta} we
take into account the different positions of the observer in the
measurement of the half-width $W_{1/2}$. From this we conclude
that observers with a line of sight perpendicular to the axis of
rotation (i.e. edge-on in the case of spiral galaxy) tend to measure
larger line widths than observers aligned with the rotation axis
(i.e. face-on). The influence of the observer position on the line
width, amounting always less than $15\%$ of a difference with respect
to the result that takes into account all the outgoing photons with
the same weight regardless of the relative observer position.

%Both for central and homogeneous distribution of sources, We found a
%dependency of the width with the rotation velocity of the gas cloud,
%this results are summarize in Figure ~\ref{figure:width}. As velocity
%increase also the equivalent width increase (for both homogeneous and
%central distribSi Limonution), it means that some photons escape with fewer
%scatterings than the static case but at the same time a few photons escape with% more scatterings, this can be seen clearly in
%Figure. (Nscatt)\\

%We also found a dependency with the viewing angles, in particular as
%the angle incre ase the width also increase Figure
%~\ref{fig:withvstheta}, this was also found in Verhemme et al. 2012
%for a rotating disk but the pure effect of rotation can not been study
%as it is in our case. Following the convention of Verhamme et al 2012
%we define $\mu=cos(\theta)$, and we can make a polynomial fit for the
%$FWHM(\mu)$ for our models, this is resumed in Table ~\ref{table:fits}

\begin{figure}
    \includegraphics[width=0.45\textwidth]{maximumvsVmax.png}
\caption{Position of the maxima
    in the outgoing spectra for differents Rotational velocities,
    (up)Central Distriution, (Down) Homogeneous Distribution.\label{fig:maximumsvsvelocity}} 
\end{figure}

The second feature in the line that we use to quantify the effect of
rotation is the position of the line maxima. These provide information
on the wavelength of the majority of the outgoing photons after
they interact with the neutral hydrogen atoms in the gas cloud. If
most of the photon escapes with a low number of scattering, its
outgoing frequency will be close to its initial frequency, that is in
the center of the line. On the contrary if the number of scatterings
is large for the average photon, its outgoing frequency will be far
from the line center. Such reasoning can be made more quantitative to
understand the dependence of the peak maxima as a function of the
hydrogen optical depth in the cloud [citation needed].

In Fig.~\ref{fig:maximumsvsvelocity} we present the position
of the maxima as a function of $V_{\rm max}$. For the photons emitted
in the line center we do not find any variation in the position of the
maxima in the range of explored parameter space (dash lines). However in the
homogeneous case we can see how the maxima goes to $x_{m}=0$, meaning
that the double peak is converted into a single peak.

This transition to a single peak line occurs for the systems where it
becomes easier for a bulk of the photons to escape with the lowest
number of scatterings possible. This can explain how the single peak
stage can be achieved in the homogeneous source distribution where
there is fraction of the photons inside a photosphere region with
$\tau_{H}\sim 1$ which allows them to escape within one
scatter. Increasing the rotational velocity makes it easier for the
photons in this photosphere region to escape.

We also take into account the viewing angle $\mu$ for all the optical 
depths and rotational velocities and the two different photon distribution.
But we don't find any dependency meaning that the peaks would be placed in
the same place no matter the position of an external observer.

Finally we also report on the effect of the neutral Hydrogen
optical depth $\tau_{H}$ on the maxima position $x_{m}$
Fig.~\ref{fig:maximumsvsvelocity}. We find that, at fixed
rotation velocity, the position of the maxima increases with optical
depth as expected from basic theoretical considerations. We compare
our results with the expected theoretical scaling for an infinite
slab.

>>>>

\newcommand{\LCDM}{$\Lambda$CDM~}
\newcommand{\beq}{\begin{eqnarray}}  
\newcommand{\eeq}{\end{eqnarray}}  
\newcommand{\zz}{$z\sim 3$} 
\newcommand{\apj}{ApJ}  
\newcommand{\apjs}{ApJS}  
\newcommand{\apjl}{ApJL}  
\newcommand{\aj}{AJ}  
\newcommand{\mnras}{MNRAS}  
\newcommand{\mnrassub}{MNRAS accepted}  
\newcommand{\aap}{A\&A}  
\newcommand{\aaps}{A\&AS}  
\newcommand{\araa}{ARA\&A}  
\newcommand{\nat}{Nature}  
\newcommand{\physrep}{PhR}
\newcommand{\pasp}{PASP}    
\newcommand{\pasj}{PASJ}    
\newcommand{\avg}[1]{\langle{#1}\rangle}  
\newcommand{\ly}{{\ifmmode{{\rm Ly}\alpha~}\else{Ly$\alpha$~}\fi}}
\newcommand{\hMpc}{{\ifmmode{h^{-1}{\rm Mpc}}\else{$h^{-1}$Mpc }\fi}}  
\newcommand{\hGpc}{{\ifmmode{h^{-1}{\rm Gpc}}\else{$h^{-1}$Gpc }\fi}}  
\newcommand{\hmpc}{{\ifmmode{h^{-1}{\rm Mpc}}\else{$h^{-1}$Mpc }\fi}}  
\newcommand{\hkpc}{{\ifmmode{h^{-1}{\rm kpc}}\else{$h^{-1}$kpc }\fi}}  
\newcommand{\hMsun}{{\ifmmode{h^{-1}{\rm {M_{\odot}}}}\else{$h^{-1}{\rm{M_{\odot}}}$}\fi}}  
\newcommand{\hmsun}{{\ifmmode{h^{-1}{\rm {M_{\odot}}}}\else{$h^{-1}{\rm{M_{\odot}}}$}\fi}}  
\newcommand{\Msun}{{\ifmmode{{\rm {M_{\odot}}}}\else{${\rm{M_{\odot}}}$}\fi}}  
\newcommand{\msun}{{\ifmmode{{\rm {M_{\odot}}}}\else{${\rm{M_{\odot}}}$}\fi}}  
\newcommand{\lya}{{Lyman$\alpha$~}}
\newcommand{\clara}{{\texttt{CLARA}}~}
\newcommand{\rand}{{\ifmmode{{\mathcal{R}}}\else{${\mathcal{R}}$ }\fi}}  
\newcommand{\hs}{{\hspace{1mm}}}  
\newcommand{\kms}{{\ifmmode{{\mathrm{\,km\ s}^{-1}}}\else{\,km~s$^{-1}$}\fi}}

@@

\begin{figure*}
  \includegraphics[width=0.45\textwidth]{SpectraDifVelocitiesCentral.png}
  \includegraphics[width=0.45\textwidth]{SpectraDifVelocitiesHOM.png}
\caption{Shape of the \ly line for
    different velocities rotational velocities for spherical
    distributions with $\tau_{H}=10^{7}$. The left (right) panel shows the central
    (homogeneous) photon distribution. All photons were taken into
    account regardless of their propagation direction.
  \label{fig:differentvelocities}}
\end{figure*}

@@

In the following subsections we characterize the line morphology by
the half-width at half intensity and the peak maxima. In order to
intepret the morphological changes in the line we also report the
median number of scatter for each \ly photon in the
simulation. Finally we measure the bulk escape fraction as a function
of rotational velocity in the presence of dust.

All the results in this section are constructed by taking into
account all outgoing photons regardless of the direction of
propagation. In the next section \ref{subsec:direction} we quantify 
the changes in the observed spectra for observers with different
viewing angles.

@@

The first quantitative conclusion of the effect of rotation in the
\ly line is that double peaks broaden and reduce their intensity,
while the line center rises. This procudes the impresion that, as the
rotational velocity increases, the double peaks are merged into a
single broad emission peak. This is msot evident in the highest
rotational velocities for the homogeneously distributed sources
(right panel in Fig.~\ref{fig:differentvelocities}).

To quantify the line broadening we measure a modified version of the full
width at half maximum (FWHM). Instead we measure it only for half of
the line, $W_{1/2}$. This definition allows us to quantify the line
width both in the cases of double and single peak emission.  In the
case of double peaked emission $W_{1/2}$ corresponds to the width of a
single peak, while in the extreme case of high rotational velocities,
when the double peak is erased, it simply correspond to half FWHM.

Figure \ref{fig:widthvsvelocity} shows how $W_{1/2}$ increases with
rotational velocity. Continous (dashed) lines connect the results for
homogeneous (central) source distribution. The line width increases
with rotational velocity up to factors of $\sim 2$ at $300$\kms with
respect to the static case. For the temperature $T=10^4$K used in our
radiative transfer calculations calculations the thermal velocity is
$v_{th}=12.8$\kms. For a model with $\tau_{H}$ it means that the
half-width can increase up to $350$\kms (at $V_{\rm max}=300$\kms)
compared to a half-width of $150$\kms in the static case.

\begin{figure}
    \includegraphics[width=0.45\textwidth]{WidthvsVmax.png}
    \caption{Half-width for the non-dusty models as a function of
      rotational velocity $V_{\rm max}$. Continous (dashed) lines
      correspond to homogeneous (central) source
      distributions. \label{fig:widthvsvelocity}} 
\end{figure}

\begin{figure}
    \includegraphics[width=0.45\textwidth]{maximumvsVmax.png}
\caption{Position of the peak maxima as a function of rotational
  velocity $V_{\rm max}$. Continous (dashed) lines correspond to
  homogeneous (central) source distributions. A value of $x_{\rm
    max}=0$ indicates that line becomes single
  peaked. \label{fig:maximumsvsvelocity}}  
\end{figure}

Figure \ref{fig:maximumsvsvelocity} shows the position for the peak
maxima as a function of the rotational velocity $V_{\rm max}$. This
figure shows us clearly that in the case of central distributed
sources there is barely any change with rotational velocity, at least
in the range of explirea parameters. However, in the case of
homogeneously emitted sources the maxima position remain close to
constant until beyond some velocity threshold the line becomes single
peaked with $x_{\rm max}=0$.

The transition to a single peak line most probably occurs for systems
where it  becomes easier for the bulk of the photons to escape with the lowest
number of scatterings possible, allowing them not to move very far
from the center of the line. This hypothesis can also help to explain how the
single peak stage is easily achieved in the homogeneous source
distribution compared to the central distribution. In the homogeneous
case there is a fraction of the photons that are inside a photosphere region with
$\tau_{H,r}\ll \tau_{H}$ where $\tau_{H,r}$ is the optical depth from
the radius of emission to the sphere's surface. This conditions allows
the photons to escape with much less scatterings compared to the
photons emitted at the very center of the sphere. In turn, it gives
the photons less scatterings to be placed far from the line
center. Increasing the rotational velocity $V_{\rm max}$ reduces the
optical depth making the  photosphere region effectively larger,
increasing the number of photons escaping close to the lines's center.

For larger values of $\tau_{H}$ the transitional velocity $V_{\rm
  trans}$ is also larger. In our models we find the following
correspondance between the optical depth $\tau_{\rm H}=\{10^5, 10^6,
10^7\}$ and the transitional velocities $V_{\rm
  trans}=\{50-100\kms, 100-200\kms, 200-300\kms\}$ which can only be
constrained to be in the range of velocities in the models that gave a $x_{\rm
  max}\neq 0$ and $x_{\rm max}=0$.
